Based on polling as of 27 October 2024, Harris expects to win on 50%, leads Trump by about 3% in 2024 US election

### 2024-10-30 paper/paper.qmd

library(tidyverse)
```

@@

## Outcome variables

@@

@@

```{r}
model_harris_weighted <- lm(pct ~ national_poll + pollster + population, data = harris_data, weights = combined_weight)
```

@@

```
@fig-national-support-trendsillustrates illustrates the polling support trends for Kamala Harris and Donald Trump across the country from August to October. The scatter points represent individual poll results, with blue for Harris and red for Trump, while each dot's size reflects the weight of the poll, accounting for factors like sample size, quality, and recency. Harris’s support trend line generally remains above Trump’s, with an average support level around 49%, compared to Trump’s approximate average of 47% over this period. Toward the end of October, both candidates show a slight increase, with Harris’s support reaching just over 50% and Trump’s approaching 49%. This trend suggests a potential narrowing in support levels as the election approaches, providing a quantitative snapshot of national polling dynamics.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false
set.seed(233)
new_data_harris <- data.frame(
  end_date = seq(
    min(harris_data$end_date),
    max(harris_data$end_date),
    length.out = 1000 
  ),
  national_poll = rep(c(0, 1), length(1000)),
  pollster = factor(sample(levels(harris_data$pollster), size = 1000, replace = TRUE), 
                    levels = levels(harris_data$pollster)),  
  population = factor(sample(c("lv","rv"), size = 1000, replace = TRUE)),
  sample_size = sample(1:2300, size = 1000, replace = TRUE),
  pollster_quality= sample(2:4, size = 1000, replace = TRUE)
)|>
  mutate(
    # Sample size weight (assuming a maximum of 2300 responses)
    sample_size_weight = pmin(sample_size / 2300, 1),  # Set sample size; adjust accordingly

@@

    # Recency weight (this can be based on the date difference from a certain cutoff, e.g., "2024-11-05")
    recency = as.numeric(difftime( as.Date("2024-11-05"),as.Date(end_date), units = "days")),  # Adjust for the sequence
    recency_weight = exp(-recency * 0.1),  # Adjust decay rate as needed
        recent_poll_count = sum(recency < 30),
    poll_frequency_weight = ifelse(recent_poll_count > 4, 4 / recent_poll_count, 1),
    # Combine all weights into a final combined weight
    combined_weight = recency_weight * sample_size_weight * poll_frequency_weight * pollster_quality_weight,
    candidate="Harris"
  )
new_data_trump <- data.frame(
  end_date = seq(
    min(trump_data$end_date),
    max(trump_data$end_date),
    length.out = 1000
  ),
  pollster = factor(sample(levels(trump_data$pollster), size = 1000, replace = TRUE), 
                    levels = levels(trump_data$pollster)),  
  population = factor(sample(c("lv","rv"), size = 1000, replace = TRUE)),
   national_poll = rep(c(0, 1), length(1000)),  
  sample_size =sample(1:2300, size = 1000, replace = TRUE),
  pollster_quality= sample(2:4, size = 1000, replace = TRUE)
)|>
  mutate(
    # Sample size weight (assuming a maximum of 2300 responses)
    sample_size_weight = pmin(sample_size / 2300, 1),  # Set sample size; adjust accordingly

    # Recency weight (this can be based on the date difference from a certain cutoff, e.g., "2024-11-05")
    recency = as.numeric(difftime( as.Date("2024-11-05"),as.Date(end_date), units = "days")),  # Adjust for the sequence
    recency_weight = exp(-recency * 0.1),  # Adjust decay rate as needed

    # Poll frequency weight (you can set a condition based on a hypothetical recent poll count)
    recent_poll_count = sum(recency < 30),
    poll_frequency_weight = ifelse(recent_poll_count > 4, 4 / recent_poll_count, 1),

    # Combine all weights into a final combined weight
    combined_weight = recency_weight * sample_size_weight * poll_frequency_weight * pollster_quality_weight,
    candidate="Trump"
  )
```

@@

## Key Findings and Real-World Implications {#sec-key-findings}
Our model reveals a portrait of a nation deeply divided, with Harris holding a narrow 49% edge over Trump’s 47% in national support. However, as recent elections have shown, the national popular vote does not always translate into an Electoral College victory. The U.S. electoral system, which awards each state’s electoral votes to the candidate with the majority of votes in that state, means that winning the popular vote nationally might still leave Harris short of the presidency. This structural quirk played a defining role in the 2016 election, where Hillary Clinton’s popular vote win failed to deliver the electoral majority, paving Trump’s path to the White House. Our findings underscore this enduring tension: while Harris may have a slight national advantage, the true battle will be fought state by state, in a handful of battlegrounds that could flip the election either way.

@@

# Additional data details

@@

## Model validation

```{r}
# Set seed for reproducibility
set.seed(123)

# Sample indices for training and testing sets
train_index <- sample(seq_len(nrow(data)), size = 0.8 * nrow(data))

# Split the data into training and testing sets
train_data <- data[train_index, ]
test_data <- data[-train_index, ]

# Fit the weighted linear regression model
first_model_test <- lm(pct ~ national_poll + pollster + population, 
                     data = data, 
                  data = train_data)

# Predict on test data
predictions <- predict(first_model_test, newdata = test_data)

# Calculate RMSE using the correct outcome variable
# Assuming 'pct' is the actual observed outcome variable in test_data
rmse <- sqrt(mean((predictions - test_data$pct)^2, na.rm = TRUE))

# Print RMSE
print(paste("RMSE:", rmse))

@@

@@

@fig-stanareyouokay-1 is a residual plot of un-weighted model. It shows... This suggests...

@fig-stanareyouokay-2 is a Q-Q plot of un-weighted model plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2
#residual plot
plot(first_model, which = 1,main ="Residual Plot of unweighted linear regression model residualsmodel" )
# Q-Q plot for the residuals of the model
qqnorm(residuals(first_model),main= "Q-Q Plot of unweighted linear regression model residuals
")
qqline(residuals(first_model), col = "red")

@@

@@

In essence, statistics is about collecting data and making informed
conclusions, even though we can never access every piece of
information.\
From @tellingstories, we defined three key terms as:\
Target population : The collection of all items about which we would
like to speak. / The entire group about which we want to draw
conclusions\
Sampling frame : A list of all the items from the target population that
we could get data about.\
Sample : The items from the sampling frame that we get data about.\

>>>>

library(tidyverse)
library(arrow)
library(dplyr)
```

```{r}
#| include: false
#| warning: false
#| message: false
harris_data <- read_parquet("~/2024 US Election Forecast/data/02-analysis_data/analysis_data_Harris.parquet")
trump_data <- read_parquet("~/2024 US Election Forecast/data/02-analysis_data/analysis_data_Trump.parquet")

@@

## Outcome variables (Estimand)

@@

The validation and diagnostic details of the model are provided in
[Appendix B](#sec-model-details)

@@

```{r}
#| include: false
#| warning: false
#| message: false
#### Calculate Weights ####
# Calculate weights for Harris data
harris_data <- harris_data |>
  mutate(
    sample_size_weight = pmin(sample_size / 2300, 1),
    pollster_quality_weight = numeric_grade / 4,
    recency_weight = exp(-recency * 0.1)
  ) |>
  group_by(pollster) |>
  mutate(
    recent_poll_count = sum(recency < 30),
    poll_frequency_weight = ifelse(recent_poll_count > 4, 4 / recent_poll_count, 1)
  ) |>
  ungroup() |>
  mutate(combined_weight = recency_weight * sample_size_weight * poll_frequency_weight * pollster_quality_weight)

# Calculate weights for Trump data
trump_data <- trump_data |>
  mutate(
    sample_size_weight = pmin(sample_size / 2300, 1),
    pollster_quality_weight = numeric_grade / 4,
    recency_weight = exp(-recency * 0.1)
  ) |>
  group_by(pollster) |>
  mutate(
    recent_poll_count = sum(recency < 30),
    poll_frequency_weight = ifelse(recent_poll_count > 4, 4 / recent_poll_count, 1)
  ) |>
  ungroup() |>
  mutate(combined_weight = recency_weight * sample_size_weight * poll_frequency_weight * pollster_quality_weight)
```

@@

```
@fig-national-support-trends illustrates the polling support trends for Kamala Harris and Donald Trump across the country from August to October. The scatter points represent individual poll results, with blue for Harris and red for Trump, while each dot's size reflects the weight of the poll, accounting for factors like sample size, quality, and recency. Harris’s support trend line generally remains above Trump’s, with an average support level around 49%, compared to Trump’s approximate average of 47% over this period. Toward the end of October, both candidates show a slight increase, with Harris’s support reaching just over 50% and Trump’s approaching 49%. This trend suggests a potential narrowing in support levels as the election approaches, providing a quantitative snapshot of national polling dynamics.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| include: false
set.seed(233)
new_data_harris <- data.frame(
  end_date = seq(
    min(harris_data$end_date),
    max(harris_data$end_date),
    length.out = 1000 
  ),
  national_poll = rep(c(0, 1), length(1000)),
  pollster = factor(sample(levels(harris_data$pollster), size = 1000, replace = TRUE), 
                    levels = levels(harris_data$pollster)),  
  population = factor(sample(c("lv","rv"), size = 1000, replace = TRUE)),
  sample_size = sample(1:2300, size = 1000, replace = TRUE),
  pollster_quality= sample(2:4, size = 1000, replace = TRUE)
)|>
  group_by(pollster)|>
  mutate(recency = as.numeric(difftime( as.Date("2024-11-05"),as.Date(end_date), units = "days")),
    recent_poll_count = sum(recency < 30),
    poll_frequency_weight = ifelse(recent_poll_count > 4, 4 / recent_poll_count, 1)
  ) |>ungroup()|>
  mutate(
    # Sample size weight (assuming a maximum of 2300 responses)
    sample_size_weight = pmin(sample_size / 2300, 1),  # Set sample size; adjust accordingly

@@

    recency_weight = exp(-recency * 0.1),  # Adjust decay rate as needed
    # Combine all weights into a final combined weight
    combined_weight = recency_weight * sample_size_weight * poll_frequency_weight * pollster_quality_weight,
    candidate="Harris"
  )
#######
new_data_trump <- data.frame(
  end_date = seq(
    min(trump_data$end_date),
    max(trump_data$end_date),
    length.out = 1000
  ),
  pollster = factor(sample(levels(trump_data$pollster), size = 1000, replace = TRUE), 
                    levels = levels(trump_data$pollster)),  
  population = factor(sample(c("lv","rv"), size = 1000, replace = TRUE)),
   national_poll = rep(c(0, 1), length(1000)),  
  sample_size =sample(1:2300, size = 1000, replace = TRUE),
  pollster_quality= sample(2:4, size = 1000, replace = TRUE)
)|>
  group_by(pollster)|>
  mutate(recency = as.numeric(difftime( as.Date("2024-11-05"),as.Date(end_date), units = "days")),
    recent_poll_count = sum(recency < 30),
    poll_frequency_weight = ifelse(recent_poll_count > 4, 4 / recent_poll_count, 1)
  ) |>
  ungroup() |>
  mutate(
    # Sample size weight (assuming a maximum of 2300 responses)
    sample_size_weight = pmin(sample_size / 2300, 1), 
    pollster_quality_weight = pollster_quality / 4,  
    recency_weight = exp(-recency * 0.1),  
    combined_weight = recency_weight * sample_size_weight * poll_frequency_weight * pollster_quality_weight,
    candidate="Trump"
  )
```

@@

## Key Findings and Real-World Implications {#sec-key-findings}

Our model reveals a portrait of a nation deeply divided, with Harris
holding a narrow 49% edge over Trump’s 47% in national support. However,
as recent elections have shown, the national popular vote does not
always translate into an Electoral College victory. The U.S. electoral
system, which awards each state’s electoral votes to the candidate with
the majority of votes in that state, means that winning the popular vote
nationally might still leave Harris short of the presidency. This
structural quirk played a defining role in the 2016 election, where
Hillary Clinton’s popular vote win failed to deliver the electoral
majority, paving Trump’s path to the White House. Our findings
underscore this enduring tension: while Harris may have a slight
national advantage, the true battle will be fought state by state, in a
handful of battlegrounds that could flip the election either way.

@@

# Additional data details {#first-model-details}

@@

## Model validation: K-Fold Cross-Validation

```{r}
#| include: false
#| message: false
#| warning: false
# Load necessary libraries
library(boot)    # For bootstrapping
library(caret)   # For cross-validation

# Set up 10-fold cross-validation
train_control <- trainControl(method = "cv", number = 10)
# Train the model
model_formula <- pct ~ national_poll + pollster + population
model_harris_cv <- train(model_formula, data = harris_data, method = "lm", trControl = train_control)
# Set up 10-fold cross-validation
train_control <- trainControl(method = "cv", number = 10)
# Train the model
model_harris_cv <- train(model_formula, data = harris_data, method = "lm", trControl = train_control)
# Print the model results

model_trump_cv <- train(model_formula, data = trump_data, method = "lm", trControl = train_control)
# Set up 10-fold cross-validation
train_control <- trainControl(method = "cv", number = 10)
# Train the model
model_trump_cv <- train(model_formula, data = trump_data, method = "lm", trControl = train_control)
# Print the model results

@@

```{r}
#| echo: false
#| message: false
#| warning: false
# Extract RMSE, R-squared, and MAE for Harris model
harris_rmse <- model_harris_cv$results$RMSE
harris_r2 <- model_harris_cv$results$Rsquared
harris_mae <- model_harris_cv$results$MAE

# Print results for Harris model
cat(sprintf("For the Harris model, the RMSE is %.2f, R-squared is %.2f, and MAE is %.2f.\n", 
            harris_rmse, harris_r2, harris_mae))

# Extract RMSE, R-squared, and MAE for Trump model
trump_rmse <- model_trump_cv$results$RMSE
trump_r2 <- model_trump_cv$results$Rsquared
trump_mae <- model_trump_cv$results$MAE

# Print results for Trump model
cat(sprintf("For the Trump model, the RMSE is %.2f, R-squared is %.2f, and MAE is %.2f.\n", 
            trump_rmse, trump_r2, trump_mae))

We use a 10-fold cross-validation on two linear regression models—one
for Harris and one for Trump. The models use three predictors:
national_poll, pollster, and population. The output provides key
metrics, which breaks down here:

RMSE (Root Mean Square Error): Measures the average magnitude of
prediction errors (lower is better).

Harris model: RMSE of 3.44, indicating an average prediction error of
around 3.44 percentage points.\
Trump model: RMSE of 4.40, showing a slightly higher prediction error on
average.\

R-squared: Represents the proportion of the variance in the response
variable explained by the model (higher is better).\
Harris model: R-squared of 0.34, meaning the model explains about 34.6%
of the variance in Harris's polling data.\
Trump model: R-squared of 0.34 as well, indicating similar explanatory
power for Trump's polling data.

MAE (Mean Absolute Error): Shows the average absolute difference between
observed and predicted values (lower is better).\
Harris model: MAE of 2.48, meaning that, on average, the predictions are
off by 2.48 percentage points.\
Trump model: MAE of 3.15, indicating slightly less precise predictions
compared to the Harris model.\

Interpretation Summary Predictive Accuracy: The Harris model has
slightly better predictive accuracy than the Trump model, as reflected
by its lower RMSE and MAE values.\
Model Fit: Both models explain roughly 34% of the variance in their
respective datasets. This suggests that other factors not included in
the model may play a significant role in explaining the remaining
variance.\
This summary indicates the models are moderately predictive, with room
for improvement in accuracy and fit, potentially by adding more
predictors or adjusting model specifications.\

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Diagnostics of model using residual vs fitted plot and norm Q-Q plot\n-Support for Harris"
#| fig-subcap: ["Residual Plot for Unweighted Model", "Q-Q Plot for Unweighted Model", "Residual Plot for Weighted Model", "Q-Q Plot for Weighted Model"]
#| layout-ncol: 2

# Residual plot for unweighted model
plot(harris_unweighted_model, which = 1, main = "Residual Plot of Unweighted Linear Regression Model")

# Q-Q plot for the residuals of the unweighted model
qqnorm(residuals(harris_unweighted_model), main = "Q-Q Plot of Unweighted Linear Regression Model Residuals")
qqline(residuals(harris_unweighted_model), col = "red")

# Residual plot for weighted model
plot(harris_weighted_model, which = 1, main = "Residual Plot of Weighted Linear Regression Model")

# Q-Q plot for the residuals of the weighted model
qqnorm(residuals(harris_weighted_model), main = "Q-Q Plot of Weighted Linear Regression Model Residuals")
qqline(residuals(harris_weighted_model), col = "red")

@@

Generally, we use Residual vs Fitted plot and Q-Q Plot diagnostic our
model. Residual vs Fitted plot aare Residuals (differences between
observed and predicted values) plotted against fitted values. Ideally,
these residuals should be randomly scattered around the zero line to
indicate that the model does not have systematic errors. The Q-Q plot
for the unweighted model shows how the residuals align with a
theoretical normal distribution. Ideally, residuals should follow a
straight line in this plot if they are normally distributed, which is an
assumption of linear regression.

@fig-stanareyouokay-1 is a residual plot of un-weighted model for Harris
support. It showsthe residuals are generally spread around zero, with no
clear pattern. This suggests the model is relatively well-specified.
However, there is a slight curvature, indicating potential non-linearity
that the model may not fully capture. A few notable outliers with larger
residuals might be influencing the model, indicating that some data
points have more significant prediction errors.

@fig-stanareyouokay-2 is a Q-Q plot of un-weighted model plot for Harris
support. It shows most residuals fall along the line, especially in the
middle range. This suggests that our model satisfy the normality
assumption. However, some points at the tails deviate, indicating
potential outliers or non-normality in the extreme residual values. This
slight deviation at the ends suggests the model might have some issues
with extreme predictions but performs reasonably well overall.

@fig-stanareyouokay-3 is a residual plot of weighted model for Harris
support.It shows residuals are again plotted against fitted values.
Similar to the unweighted model, the residuals are mostly centered
around zero, indicating that the weighted model captures the general
trend without significant systematic bias. The curvature is slightly
reduced compared to the un-weighted model, suggesting that weighting has
helped in addressing some of the non-linearity observed in the
un-weighted model. However, some residuals are still notably large,
which may indicate outliers that influence the model despite the
weighting scheme. This suggests that while the weighted model performs
better in terms of capturing non-linearity, further refinement might
still be beneficial.

@fig-stanareyouokay-4 is a Q-Q Plot of weighted model plot for Harris
support.The residuals generally align with the theoretical normal
distribution line, particularly in the central range, indicating that
the residuals of the weighted model are close to normal. Similar to the
unweighted model, there are deviations at the tails, though they appear
less pronounced. This suggests that the weighting scheme has slightly
improved the distribution of residuals, making the model’s predictions
more robust. However, some extreme values remain, which could still
affect model

In summary, both models show a reasonably good fit, with the weighted
model offering slight improvements in handling non-linearity and extreme
values. However, both models exhibit minor deviations from normality and
a few notable outliers, which may warrant further model adjustments for
improved prediction accuracy.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-dia
#| fig-cap: "Diagnostics of model using residual vs fitted plot and norm Q-Q plot\n-Support for Trump"
#| fig-subcap: ["Residual Plot for Unweighted Model", "Q-Q Plot for Unweighted Model", "Residual Plot for Weighted Model", "Q-Q Plot for Weighted Model"]
#| layout-ncol: 2

# Residual plot for unweighted model
plot(trump_unweighted_model, which = 1, main = "Residual Plot of Unweighted Linear Regression Model")

# Q-Q plot for the residuals of the unweighted model
qqnorm(residuals(trump_unweighted_model), main = "Q-Q Plot of Unweighted Linear Regression Model Residuals")
qqline(residuals(trump_unweighted_model), col = "red")

# Residual plot for weighted model
plot(trump_weighted_model, which = 1, main = "Residual Plot of Weighted Linear Regression Model")

# Q-Q plot for the residuals of the weighted model
qqnorm(residuals(trump_weighted_model), main = "Q-Q Plot of Weighted Linear Regression Model Residuals")
qqline(residuals(trump_weighted_model), col = "red")

@fig-dia-1 shows the residuals plotted against the fitted values for the
unweighted model. It shows that the residuals are generally spread
around zero with no clear pattern, suggesting that the model is
relatively well-specified. However, there is a slight curvature,
indicating possible non-linearity that the model may not fully capture.
Some notable outliers with larger residuals suggest that certain data
points have significant prediction errors, potentially influencing the
model.

@fig-dia-2 shows the Q-Q plot for the unweighted model, showing that
most residuals fall along the line, especially in the middle range,
suggesting that the model satisfies the normality assumption. However,
some points at the tails deviate, indicating potential outliers or
non-normality in extreme residual values. This deviation at the ends
suggests the model may face issues with extreme predictions, though it
performs reasonably well overall.

@fig-dia-3 shows residuals plotted against fitted values. Similar to the
unweighted model, the residuals are centered around zero, indicating
that the weighted model captures the overall trend without significant
systematic bias. The slight curvature seen in the unweighted model is
reduced here, suggesting that weighting has addressed some of the
non-linearity. However, some large residuals remain, which could
indicate outliers that affect the model even with the weighting scheme.
This suggests that while the weighted model has improved in handling
non-linearity, further refinements could enhance accuracy.

@fig-dia-4 shows the Q-Q plot for the weighted model compares residuals
with a theoretical normal distribution. Here, the residuals generally
align with the theoretical line, especially in the central range,
indicating that the residuals of the weighted model are close to normal.
Similar to the unweighted model, some deviations occur at the tails,
though they appear less pronounced, suggesting that the weighting scheme
has slightly improved the normality of residuals. Nonetheless, some
extreme values persist, which may impact model robustness in cases of
outliers.

Summary Both models exhibit a reasonably good fit, with the weighted
model offering slight improvements in managing non-linearity and extreme
values. Despite this, both models show minor deviations from normality
and some notable outliers, suggesting that further model adjustments may
be beneficial for improved prediction accuracy.

@@

In essence, statistics is about collecting data and making informed
conclusions, even though we can never access every piece of
information.\
From @tellingstories, we defined three key terms as:\
Target population : The collection of all items about which we would
like to speak/ the entire group about which we want to draw.
conclusions\
Sampling frame : A list of all the items from the target population that
we could get data about.\
Sample : The items from the sampling frame that we get data about.\
